Relative parabolic factorizations of Chevalley groups

### 2015-09-09 dennis-vaserstein.tex
Slightly improved the exposition

Let $\Phi$ be an irreducible root system of rank $\ell$.
In what follows, $r$ and $s$ are two distinct integers such that $1 \leq r, s \leq \ell$.
From Levi decomposition~\eqref{rel:Levi-decomp} it follows that the following four subsets of $\E(\Phi, R, I)$ are equal:
\begin{multline*}
\U(\Phi^+, I)\cdot \U(\Phi^-, I) \cdot \E(\Delta_r, R, I) \cdot \EP_s(R, I) = \\
= \U(\Sigma_r, I)\cdot \U(\Sigma^-_r, I) \cdot \E(\Delta_r, R, I) \cdot \EP_s(R, I) = \hspace{5em} \\
\hspace{5em} = \EP_r(R, I) \cdot \E(\Delta_s, R, I) \cdot \U(\Sigma_s^-, I)\cdot \U(\Sigma_s, I) = \\
= \EP_r(R, I) \cdot \U(\Sigma^-_r \cap \Sigma^-_s, I) \cdot \EP_s(R, I).
\end{multline*}

\begin{dfn}
 Denote by $A_{rs}$ any of the above equal subsets.
 We call the decomposition of the form $\E(\Phi, R, I) = A_{rs}$ the {\it relative Dennis---Vaserstein} decomposition.
\end{dfn}

The main result of this subsection is the following theorem which is a relative version of \cite[Lemma~2.1]{ST76} and \cite[Theorem~2.5]{St78}).
\begin{thm}\label{thm:DennisVaserstein}
The relative elementary subgroup $\E(\Phi, R, I)$ admits Dennis---Vaserstein decomposition under the following assumptions on $I$.
\[\begin{array}{l@{\qquad}l@{\qquad}c@{\quad}c@{\qquad}l}
   & \Phi                                 & s    & r      & \text{ring condition} \\ \hline\vphantom{\Bigl(}
1. & \rA_\ell,\ \ell\geqslant 2            & 1    & \ell   & \sr(I) \leqslant \ell-1 \\    
2. & \rB_\ell,\ \ell\geqslant 2            & 1    & \ell   & \sr(I) \leqslant \ell-1 \\
3. & \rC_\ell,\ \ell \geqslant 2          & 1    & \ell   & \sr(I) \leqslant \ell-1 \\
4. & \rD_\ell,\ \ell\geqslant 4           & 1    & \ell   & \sr(I) \leqslant \ell-2 \\ 
5. & \rD_\ell,\ \ell\geqslant 4           & \ell & \ell-1 & \sr(I) \leqslant 2      \\
6. & \rE_\ell,\ \ell=6,7                  & \ell & 2      & \sr(I) \leqslant \ell-3 \\ 
7. & \rE_\ell,\ \ell=6,7                  & \ell & 1      & \asr(I)\leqslant \ell-2 \end{array}\]
\end{thm}
Consider the following two submonoids of $\E(\Phi, R)$:
\[ N_{rs} = \left\{ g\in \E(\Phi, R)\ \middle|\ g \cdot A_{rs} \cdot g^{-1} \subseteq A_{rs} \right\}, \quad
   L_{rs}=  \left\{ g\in \E(\Phi, R, I)\ \middle|\ g \cdot A_{rs} \subseteq A_{rs} \right\}. \]

@@

Let us construct a set $B$ satisfying the assumptions of \cref{lemma:Stein_reduction} explicitly.

Denote by $\pi$ the irreducible representation of $G=\G(\Phi, R)$ on a free module $V$ with the highest weight $\varpi_s$.
The type of $\pi$ and the total number of weights in $\Lambda(\pi)$ are listed in Figure~\ref{fig:table}.

Clearly, after restricting the representation $(V, \pi)$ to the subgroup $G'=\G(\Delta_r, R)$ it becomes reducible.
Denote by $(V', \pi')$ the irreducible component corresponding to the same highest weight $\varpi_s \in \Lambda(\pi)$.

Consider the following subset of $\Lambda(\pi')$:
$$ \Gamma = \{\lambda \in \Lambda(\pi') \mid \lambda - \alpha_r \in \Lambda(\pi) \}.$$

The type of $\pi'$ and the number of elements in $\Lambda(\pi')$, $\Gamma$ are listed in Figure~\ref{fig:table}.

\begin{rem}
The sets $\Lambda(\pi')$, $\Gamma$ can be visualized using the technique of {\it weight diagrams} developed in~\cite{PSV98}.
In the simplest case of a microweight representation (i.\,e. a representation without zero weights)
the weight diagram of $(V, \pi)$ is just a graph whose vertices are in one-to-one correspondence with the elements of $\Lambda(\pi)$.
Two vertices $\lambda$, $\mu$ of the diagram are joined by a bond marked $i$ iff $\lambda-\mu = \alpha_i$.

The authors of~\cite{PSV98} put the vertex corresponding to the highest weight $\varpi_s$ of $\pi$ in the upper-left corner of the weight diagram.
Thus the subset $\Lambda(\pi') \subseteq \Lambda(\pi)$ corresponds to the upper-left connected component of the graph obtained by removing all bonds marked $r$.
In turn, the set $\Gamma$ corresponds to the vertices of $\Lambda(\pi')$ incident to the removed bonds.

The action of elementary root unipotents $x_\alpha(\xi)$ on $V$ also can be visualized using weight diagrams.

\end{rem}

\begin{figure}
\[\begin{array}{l @{\qquad} l @{\qquad} l @{\qquad} c @{\quad} c @{\qquad} l @{\qquad} l}
   & \Phi                                 & (s,r)         & |\Lambda(\pi)| & \text{type of $\pi$} & \text{type of $\pi'$}        & (|\Lambda(\pi')|, |\Gamma|)  \\ \hline\vphantom{\Bigl(}
1. & \rA_\ell,\ \ell\geqslant 2           & (1,\ell)      & \ell+1         & \text{natural}       & (\rA_{l-1}, \varpi_1)        & (\ell,1)  \\     
2. & \rB_\ell,\ \ell\geqslant 2           & (1,\ell)      & 2\ell+1        & \text{natural}       & (\rA_{l-1}, \varpi_1)        & (\ell,1) \\     
3. & \rC_\ell,\ \ell\geqslant 2           & (1,\ell)      & 2\ell          & \text{natural}       & (\rA_{l-1}, \varpi_1)        & (\ell,1)  \\
4. & \rD_\ell,\ \ell\geqslant 4           & (1,\ell)      & 2\ell          & \text{natural}       & (\rA_{l-1}, \varpi_1)        & (\ell,2)  \\ 
5. & \rD_\ell,\ \ell\geqslant 4           & (\ell,\ell-1) & 2^{\ell-1}     & \text{half-spinor}   & (\rA_{l-1}, \varpi_{\ell-1}) & (\ell,\ell-2)  \\
6. & \rE_\ell,\ \ell=6,7                  & (\ell,2)      & 27, 56         & \text{minimal}       & (\rA_{l-1}, \varpi_1)        & (\ell,3)  \\ 
7. & \rE_\ell,\ \ell=6,7                  & (\ell,1)      & 27, 56         & \text{minimal}       & (\rD_{l-1}, \varpi_1)        & (2\ell-2,\ell-1)  \end{array}\]
 \caption[Table]{List of representation involved in the proof of~\cref{thm:DennisVaserstein}.} \label{fig:table}
\end{figure}

@@

Our next step is to show that $\U(\Sigma_s, R) \subseteq N_{rs}$.
First of all, notice that by~\cref{lemma:dv-normal} we have that $X_{\alpha_s}(R) \subseteq N_{rs}$.
In cases $\Phi\neq \rB_\ell, \rC_\ell$ all roots of $\Sigma^+_s$ have the same $s$-shape and length therefore the needed statement follows from~\cref{item-trans2}.
We defer the proof in the case $\Phi=\rC_\ell$ until we prove a separate lemma (see~\cref{lemma:DVcaseCl} below).

@@

>>>>

Let $\Phi$ be an irreducible root system of rank $\ell$.
In what follows, $r$ and $s$ are two distinct integers such that $1 \leq r, s \leq \ell$.
From Levi decomposition~\eqref{rel:Levi-decomp} it follows that the following four subsets of $\E(\Phi, R, I)$ are equal:
\begin{multline*}
\U(\Phi^+, I)\cdot \U(\Phi^-, I) \cdot \E(\Delta_r, R, I) \cdot \EP_s(R, I) = \\
= \U(\Sigma_r, I)\cdot \U(\Sigma^-_r, I) \cdot \E(\Delta_r, R, I) \cdot \EP_s(R, I) = \hspace{5em} \\
\hspace{5em} = \EP_r(R, I) \cdot \E(\Delta_s, R, I) \cdot \U(\Sigma_s^-, I)\cdot \U(\Sigma_s, I) = \\
= \EP_r(R, I) \cdot \U(\Sigma^-_r \cap \Sigma^-_s, I) \cdot \EP_s(R, I).
\end{multline*}
Denote by $A_{rs}$ any of the above equal subsets.

The main result of this subsection is the following theorem which is a relative version of \cite[Lemma~2.1]{ST76} and \cite[Theorem~2.5]{St78}).
\begin{thm}\label{thm:DennisVaserstein}
The relative elementary subgroup $\E(\Phi, R, I)$ admits Dennis---Vaserstein decomposition, i.\,e. a decompostion of the form $\E(\Phi, R, I) = A_{rs}$ under the following assumptions on $I$ and $\Phi$.
\[\begin{array}{l@{\qquad}l@{\qquad}c@{\quad}c@{\qquad}l}
   & \Phi                                 & s    & r      & \text{ring condition} \\ \hline\vphantom{\Bigl(}
1. & \rA_\ell,\ \ell\geqslant 2            & 1    & \ell   & \sr(I) \leqslant \ell-1 \\    
2. & \rB_\ell,\ \ell\geqslant 2            & 1    & \ell   & \sr(I) \leqslant \ell-1 \\
3. & \rC_\ell,\ \ell \geqslant 2          & 1    & \ell   & \sr(I) \leqslant \ell-1 \\
4. & \rD_\ell,\ \ell\geqslant 4           & 1    & \ell   & \sr(I) \leqslant \ell-2 \\ 
5. & \rD_\ell,\ \ell\geqslant 4           & \ell & \ell-1 & \sr(I) \leqslant 2      \\
6. & \rE_\ell,\ \ell=6,7                  & \ell & 2      & \sr(I) \leqslant \ell-3 \\ 
7. & \rE_\ell,\ \ell=6,7                  & \ell & 1      & \asr(I)\leqslant \ell-2 \end{array}\]
\end{thm}
Consider the following two submonoids of $\E(\Phi, R)$:
\[ N_{rs} = \left\{ g\in \E(\Phi, R)\ \middle|\ g \cdot A_{rs} \cdot g^{-1} \subseteq A_{rs} \right\}, \quad
   L_{rs}=  \left\{ g\in \E(\Phi, R, I)\ \middle|\ g \cdot A_{rs} \subseteq A_{rs} \right\}. \]

@@

The next step of the proof is to construct explicitly the set $B$ satisfying the assumptions of \cref{lemma:Stein_reduction}.
For this purpose we are going to use the technique of {\it weight diagrams} developed in~\cite{PSV98}.
In the simplest case of a microweight representation (i.\,e. a representation without zero weights)
the weight diagram of $(V, \pi)$ is just a graph whose vertices are in one-to-one correspondence with the elements of $\Lambda(\pi)$.
Two vertices $\lambda$, $\mu$ of the diagram are joined by a bond marked $i$ iff $\lambda-\mu = \alpha_i$.

Denote by $\pi$ the irreducible representation of $G=\G(\Phi, R)$ on a free module $V$ with the highest weight $\varpi_s$.
Clearly, after restricting the representation $(V, \pi)$ to the subgroup $G'=\G(\Delta_r, R)$ it becomes reducible.
Denote by $(V', \pi')$ the irreducible component corresponding to the same highest weight $\varpi_s \in \Lambda(\pi)$.
Set $\Gamma = \{\lambda \in \Lambda(\pi') \mid \lambda - \alpha_r \in \Lambda(\pi) \}.$

Using the language of weight diagrams subsets $\Lambda(\pi')$, $\Gamma$ can be visualized as follows.
After removing all bonds marked $r$ the diagram of $\pi$ splits into several connected components.
The subset $\Lambda(\pi') \subseteq \Lambda(\pi)$ corresponds to the component containg the vertex $\varpi_s$.
In turn, the set $\Gamma$ corresponds to vertices of $\Lambda(\pi')$ incident to any of the removed bonds.
%Calculating with weight diagrams we get the number of weights in $\Lambda(\pi')$ and $\Gamma$ (see~Figure~\ref{fig:table}).

Notice that $\Delta_s$ has type either $\rD_\ell$ (in the case $\Phi=\rE_\ell$, $(r,s)=(2,\ell)$) or $\rA_\ell$ (otherwise).
The number of elements in $\Lambda(\pi')$, $\Gamma$ are listed in Figure~\ref{fig:table}.
Clearly, in all cases $\pi'$ is the natural representation of $\G(\Delta_s, R)$.

Another advantage of weight diagrams is that they allow one to visualize the action of elementary root unipotents $x_\alpha(\xi)$ on $v = \sum_{\lambda \in \Lambda(\pi)} v_\lambda e_\lambda \in V$ described in~\cref{lemma:Matsumoto}.
Such description for microweight $\pi$ and $\alpha = \pm\alpha_i\in \pm\Pi$ is particularly simple: 
for every bond of the diagram joining vertices $\lambda$ and $\lambda + \alpha$ the element $x_{\alpha}(\xi)$ adds $\pm \xi \cdot v_\lambda$ to $v_{\lambda+\alpha}$.
See~\cite[Section~3.4]{PSV98} for more detailed description.

\begin{figure}
\[\begin{array}{l @{\qquad} l @{\qquad} l @{\qquad} c @{\quad} c @{\qquad} l @{\qquad} l}
   & \Phi                                 & (s,r)         & |\Lambda(\pi)| & \text{type of $\pi$} & |\Lambda(\pi')|& |\Gamma|  \\ \hline\vphantom{\Bigl(}
1. & \rA_\ell,\ \ell\geqslant 2           & (1,\ell)      & \ell+1         & \text{natural}       & \ell           & 1  \\     
2. & \rB_\ell,\ \ell\geqslant 2           & (1,\ell)      & 2\ell+1        & \text{natural}       & \ell           & 1  \\     
3. & \rC_\ell,\ \ell\geqslant 2           & (1,\ell)      & 2\ell          & \text{natural}       & \ell           & 1  \\
4. & \rD_\ell,\ \ell\geqslant 4           & (1,\ell)      & 2\ell          & \text{natural}       & \ell           & 2  \\ 
5. & \rD_\ell,\ \ell\geqslant 4           & (\ell,\ell-1) & 2^{\ell-1}     & \text{half-spinor}   & \ell           & \ell-2  \\
6. & \rE_\ell,\ \ell=6,7                  & (\ell,2)      & 27, 56         & \text{minimal}       & \ell           & 3       \\ 
7. & \rE_\ell,\ \ell=6,7                  & (\ell,1)      & 27, 56         & \text{minimal}       & 2(\ell-1)      & \ell-1  \end{array}\]
 \caption[Table]{List of representation involved in the proof of~\cref{thm:DennisVaserstein}.} \label{fig:table}
\end{figure}

@@

Our next step is to show that $\U(\Sigma_s, R) \subseteq N_{rs}$.
First of all, notice that by~\cref{lemma:dv-normal} we have that $X_{\alpha_s}(R) \subseteq N_{rs}$.
For $\Phi\neq \rB_\ell, \rC_\ell$ all roots of $\Sigma^+_s$ have the same $s$-shape and length hence the assertion follows from~\cref{item-trans2}.

@@

We defer the proof in the case $\Phi=\rC_\ell$ until we prove~\cref{lemma:DVcaseCl} below.

### 2015-09-09 paper.tex
Slightly improved the exposition

\section{Introduction}\label{sec:intro}
Let $R$ be a commutative unital ring and $\Phi$ be a reduced irreducible root system with a fixed basis of simple roots $\Pi$.
Denote by $G=\G(\Phi, R)$ the split simply-connected Chevalley group of type $\Phi$ over $R$ and by $\E(\Phi, R)$ its \emph{elementary subgroup}, 
i.\,e. the subgroup generated by root unipotents $x_\alpha(\xi)$, $\alpha\in\Phi$, $\xi\in R$ see~\cite{VP, St78, S}.
For an ideal $I \trianglelefteq R$ denote by $\G(\Phi, R, I)$ the \emph{relative Chevalley group} (also called \emph{principal congruence subgroup}) of level $I$, 
i.\,e. the kernel of the map $\G(\Phi, R)\rightarrow\G(\Phi, R/I)$ induced by the canonical projection $R\rightarrow R/I$.

The main goal of the present paper is to study parabolic factorizations of Chevalley groups and their analogues for principal congruence subgroups $\G(\Phi, R, I)$.
By a parabolic factorization we mean a decomposition of $G$ into a product of several parabolic and unipotent subgroups.
To be more specific, we study relative versions of the following three kinds of decompositions:
\begin{itemize}
\item \emph{Gauss decomposition,} i.\,e. a decomposition of the form $\E(\Phi,R)=\B \cdot \U^- \cdot \U$, where $\B$ denotes the standard Borel subgroup,
      and $\U$, $\U^-$ are the unipotent radicals of $\B$ and the opposite Borel subgroup $\B^-$;
\item \emph{Bass---Kolster decomposition} $G = \Par_i \cdot \U_i^- \cdot \U_i \cdot \U_i^-$, where 
      $\Par_i$, $\Par_i^-$ are the opposite maximal parabolic subgroups of $\G$ corresponding to $i$-th simple root of $\Phi$ and $\U_i$, $\U_i^-$ are their unipotent radicals;
\item \emph{Dennis---Vaserstein decomposition} $\E(\Phi, R) = \EP_i \cdot (\U_i^- \cap \U_j^-) \cdot \EP_j$, $i\neq j$,
      where $\EP_i$ denotes the subgroup generated by root unipotents $x_\alpha(\xi)$ contained in $\Par_i$.
\end{itemize}
%TODO: Insert some links to VIP's papers :)

@@

\begin{tabular}{ll}
If $\lambda\in\Lambda^*,\ \lambda+\alpha\notin\Lambda$, then & $x_\alpha(\xi)v^\lambda=v^\lambda$; \\
If $\lambda,\lambda+\alpha\in\Lambda^*$, then & $x_\alpha(\xi)v^\lambda=v^\lambda\pm\xi v^{\lambda+\alpha}$; \\
If $\alpha\notin\Lambda^*$, then & $x_\alpha(\xi)v^0=v^0$ for any $v^0\in V^0$; \\
If $\alpha\in\Lambda^*$, then & $x_\alpha(\xi)v^{-\alpha}=v^{-\alpha}\pm\xi v^0(\alpha)\pm\xi^2 v^\alpha$ and \\
& $x_\alpha(\xi)v^0=v^0\pm\xi\alpha_*(v^0)v^\alpha$.
\end{tabular}
% \begin{thmlist}
%  \item $x_{\alpha}(\xi) v_\lambda = v_\lambda$ for $\lambda\in\Lambda^*(\pi), \lambda+\alpha\not\in\Lambda(\pi)$;
%  \item $x_\alpha(\xi)v_\lambda=v_\lambda\pm\xi v_{\lambda+\alpha}$ for $\lambda,\lambda+\alpha\in\Lambda^*(\pi)$;
%  \item $x_\alpha(\xi)v_0 = v_0$ for every $v_0\in V^0$, $\alpha\not\in\Lambda^*(\pi)$;
%  \item $x_\alpha(\xi)v_{-\alpha}=v_{-\alpha}\pm\xi v_0(\alpha)\pm\xi^2 v_{\alpha}$,\\
%   $x_\alpha(\xi)v_0 = v_0\pm\xi\alpha_*(v_0)v_\alpha$ for $\alpha\in\Lambda^*(\pi)$.
% \end{thmlist}
\end{lemma}

>>>>

\section{Introduction}\label{sec:intro}
Let $R$ be a commutative unital ring and $\Phi$ be a reduced irreducible root system of rank $\ell$ with a fixed basis of simple roots $\Pi=\{\alpha_1,\ldots, \alpha_\ell\}$ numbered as in Bourbaki's book~\cite{Bou}.
Denote by $G=\G(\Phi, R)$ the split simply-connected Chevalley group of type $\Phi$ over $R$ and by $\E(\Phi, R)$ its \emph{elementary subgroup}, 
i.\,e. the subgroup generated by root unipotents $x_\alpha(\xi)$, $\alpha\in\Phi$, $\xi\in R$ see~\cite{VP, St78, S}.
For an ideal $I \trianglelefteq R$ denote by $\G(\Phi, R, I)$ the \emph{relative Chevalley group} (also called \emph{principal congruence subgroup}) of level $I$, 
i.\,e. the kernel of the map $\G(\Phi, R)\rightarrow\G(\Phi, R/I)$ induced by the canonical projection $R\rightarrow R/I$.

The main goal of the present paper is to study parabolic factorizations of Chevalley groups and their analogues for principal congruence subgroups $\G(\Phi, R, I)$.
By a parabolic factorization we mean a decomposition of $G$ into a product of several parabolic and unipotent subgroups.
To be more specific, we study relative analogues of the following three kinds of decompositions:
\begin{itemize}
\item \emph{Gauss decomposition,} i.\,e. a decomposition of the form $\E(\Phi,R)=\B \cdot \U^- \cdot \U$, where $\B$ denotes the standard Borel subgroup,
      and $\U$, $\U^-$ are the unipotent radicals of $\B$ and the opposite Borel subgroup $\B^-$;
\item \emph{Bass---Kolster decomposition} $G = \Par_i \cdot \U_i^- \cdot \U_i \cdot \U_i^-$, where 
      $\Par_i$, $\Par_i^-$ are the opposite maximal parabolic subgroups of $\G$ corresponding to $i$-th simple root of $\Phi$ and $\U_i$, $\U_i^-$ are their unipotent radicals;
\item \emph{Dennis---Vaserstein decomposition} $\E(\Phi, R) = \EP_i \cdot (\U_i^- \cap \U_j^-) \cdot \EP_j$, $i\neq j$,
      where $\EP_i$ denotes the subgroup generated by root unipotents $x_\alpha(\xi)$ contained in $\Par_i$.
\end{itemize}
%TODO: Insert some links to VIP's papers :)

@@

\begin{tabular}{ll}
If $\lambda\in\Lambda^*,\ \lambda+\alpha\notin\Lambda$, then & $x_\alpha(\xi)v^\lambda=v^\lambda$; \\
If $\lambda,\lambda+\alpha\in\Lambda^*$, then & $x_\alpha(\xi)v^\lambda=v^\lambda\pm\xi v^{\lambda+\alpha}$; \\
If $\alpha\notin\Lambda^*$, then & $x_\alpha(\xi)v^0=v^0$ for any $v^0\in V^0$; \\
If $\alpha\in\Lambda^*$, then & $x_\alpha(\xi)v^{-\alpha}=v^{-\alpha}\pm\xi v^0(\alpha)\pm\xi^2 v^\alpha$ and \\
& $x_\alpha(\xi)v^0=v^0\pm\xi\alpha_*(v^0)v^\alpha$.
\end{tabular}
\end{lemma}
